---
title: "Predicting Students' Chance of Admission Using Beta Regression"
author: "Marcus Chery and Keiron Green"
date: '`r Sys.Date()`'

format: 
  html:
    code-fold: true
course: STA 6257 - Advanced Statistical Modeling
bibliography: references.bib # file contains bibtex for references
#always_allow_html: true # this allows to get PDF with HTML features
self-contained: true
execute: 
  warning: false
  message: false
editor: 
  markdown: 
    wrap: 72
---

[Website](https://willg90.github.io/STA6257_Project_BetaRegression)

[Slides](https://willg90.github.io/STA6257_Project_BetaRegression/beta_slides.html)

## Introduction

The literature review delves into the development and application of
beta regression, a statistical method used for handling continuous
bounded response variables typically constrained within the open
interval (0, 1). This approach is notably significant in modeling
proportions and percentages, offering a nuanced and interpretable
framework for understanding the impact of predictor variables. The
foundational work by [@ferrari2004] in introducing beta regression has
been a pivotal point in statistical methodology, setting the stage for
numerous advancements and applications in this field.

This report delves into the application of beta regression within the
context of university admissions data analysis. Our objective is to
provide a thorough understanding of beta regression, its practical
implementation using the "betareg" package in R, and its real-world
relevance through predicting a student's chance of admission to
university given certain factors. Throughout the report, we offer
practical examples, code snippets, and graphs to facilitate hands-on
learning. Additionally, we address the limitations of beta regression
and discuss potential future research directions. The goal of this
report is to equip readers with the knowledge and skills needed to
effectively utilize beta regression in university admissions data.

Beta regression's core premise involves assuming the dependent variable
to follow a beta distribution. This assumption allows for a flexible yet
robust framework to model the relationship between dependent and
independent variables, particularly when dealing with data such as
rates, proportions, and percentages. The method's adaptability is
further enhanced by using link functions---such as logit, probit, and
log-log---facilitating the understanding of how predictor variables
influence the response variable's distribution, both in terms of its
mean and variability.

The introduction of the "betareg" package for R by [@cri2010] marked a
significant milestone, providing a practical and user-friendly tool for
implementing beta regression models. This package has been instrumental
in popularizing beta regression in data analysis, particularly in
scenarios involving continuous bounded response variables.

Subsequent research has expanded upon the initial framework of beta
regression. [@guolo2014] introduced a model incorporating Gaussian
copula to address serial dependence in time series data, broadening the
scope of beta regression to new areas such as epidemiological trend
analysis. [@espinheira2008] contributed significantly by introducing
innovative residuals based on Fisher's scoring iterative algorithm,
thereby enhancing model assessment capabilities and offering solutions
for bias correction.

Further advancements in the methodology have been made over the years.
[@simas2010] delved into the issue of bias in parameter estimation
within beta regression models, providing strategies for bias correction.
[@schmid2013] developed the concept of "boosted beta regression",
catering to complex modeling situations involving multicollinearity,
nonlinear relationships, and overdispersion. Their approach utilizes the
gamboostLSS algorithm, enabling efficient estimation of beta regression
models and demonstrating their efficacy through applications in
ecological data.

[@abonazel2022] focused on improving estimation in the presence of
multicollinearity, proposing the Dawoud--Kibria estimator as an
alternative to traditional maximum likelihood estimators. [@douma2019]
highlighted the challenges and solutions for analyzing proportional data
in ecological and evolutionary studies, advocating for beta and
Dirichlet regression techniques. [@couri2022] evaluated various
algorithms for estimating parameters in beta regression models, offering
insights into overcoming computational challenges in parameter
estimation.

Of particular note is the work of [@ospina2012], who introduced
zero-or-one inflated beta regression models. This approach is
specifically tailored for datasets containing continuous proportions
that are inflated with zeros or ones, offering a robust solution for
such mixed continuous-discrete distributions.

In synthesizing these developments, the review underscores the
versatility and continued evolution of beta regression as a statistical
tool. Its relevance across various disciplines and data types is
evident, showcasing its utility in real-world scenarios. This paper
specifically applies beta regression to predict students' chances of
admission, demonstrating the methodology's applicability in educational
data analysis and offering insights into factors influencing admission
decisions.

## Methods

Our objective is to investigate the relationship and effect of the given
variables on the response variable 'chance_of_admit'. Being that the
chosen response is proportional in nature, linear regression is not
appropriate. The suitable approach is the implementation of beta
regression. The reason as to why linear regression is not appropriate is
due to predicted values falling outside the bound of (0,1) where as all
the proportional values are within the bounds of 0 and 1.Our objective
is to implement beta regression to predict the proportion values based
the student's indicated chance of admission.

After the data has been cleaned appropriately, the beta regression
algorithm is implemented. Before delving into the implementation of the
beta regression let's briefly analyze the mathematics of the beta
regression algorithm. The beta regression is based on the logit
function. The logit function plots probabilities between the interval 0
and 1. The logit function can be seen below.
$$ logit(p) = log(p/(1-p))$$

Another formula which will need to be accounted for in predicting the
proportionality values is the beta density.The beta density is
especially useful in modelling probabilities as it works on the interval
\[0,1\] When implementing beta regression it is assumed that the
response variable follows a beta distribution hence the name beta
regression.

The beta density formula is below. $$ 
f(x;\alpha,\beta) =\frac{ \Gamma(p + q)}{\Gamma(p)*\Gamma(q)}x^{p-1}(1-x)^{q-1}
$$ , $$ 0 < y < 1 $$ where $p,q$ \> 0 and $\Gamma (.)$ is the Gamma
function. An alternate parameterization was proposed by (Ferrari and
Cribari-Neto 2004) by setting $u = p/(p+q)$ and $\phi = p + q$

$$
 f(y;\mu,\phi)=\frac{\Gamma(\phi)}{\Gamma(\mu\phi)\Gamma((1-\mu)\phi)} y^{\mu\phi-1}(1-y)^{(1-\mu)(\phi-1)}
$$

$$0<y<1$$

with $0 < u < 1$ and $\phi > 0$.

We denote $y\sim B(u,\phi)$ where $E(y)=u$ and $VAR(y)=u(1-u)/(1+\phi)$.
The $\phi$ is known as the precision parameter. For a fixed value of $u$
as the $\phi$ increases the variance of the response variable $y$
decreases.

Let $y_1,.....,y_n$ be a random sample set given that
\$y_i\simB(u_i,\phi), i = 1,......,n. The beta regression model is
defined as $$ g(u_i) = x^T_iB = n_i$$ where $B=(B_1,....,B_k)$ is a
$k*1$ vector of unknown regression parameters
$(k<n),x_i=(x_i,......x_n)^T$ is the vector of k regressors(independent
variables or covariates) and $n_i$ is a linear predictor -
i.e.,$n_i = B_1x_{i1}+...+B_kx_{ik};$ usually $x_{i1}$ = 1 for all $i$
so that the model has an intercept.The $g(.): (0,1) -> IR$ represent the
link function.

## Data Exploration

This section aims to identify key variables for constructing a beta
regression model to predict admission chances. Initially, we analyze the
dataset to determine the correlation between various predictors and the
chance of admission. Following this, we use these insights to develop
and evaluate several models, focusing on their predictive accuracy and
effectiveness. This process is critical for establishing a reliable
model that accurately forecasts admission probabilities based on the
identified significant predictors.

### Data and Visualization

Data: [**University Admission
Data**](https://www.kaggle.com/datasets/akshaydattatraykhare/data-for-admission-in-the-university/data)

### Attribute Information

+-------------+-------------+-------------+-------------------+
| Variable    | Parameter   | Range       | Description       |
+=============+=============+:============+===================+
| GRE Scores  | gre_score   | 290 - 340   | Quantifies a      |
|             |             |             | candidate's       |
|             |             | (340 scale) | performance on    |
|             |             |             | the Graduate      |
|             |             |             | Record            |
|             |             |             | Examination, with |
|             |             |             | a maximum score   |
|             |             |             | of 340            |
+-------------+-------------+-------------+-------------------+
| TOEFL       | to          | 92 - 120    | Measures English  |
| Scores      | efl_score   |             | language          |
|             |             | (120 scale) | proficiency,      |
|             |             |             | scored out of a   |
|             |             |             | total of 120      |
|             |             |             | points            |
+-------------+-------------+-------------+-------------------+
| Un iversity | universi    | 1 to 5 with | Rates             |
| Rating      | ty_rating   | 5 being the | universities on a |
|             |             | highest     | scale from 1 to   |
|             |             | rating      | 5, indicating     |
|             |             |             | their overall     |
|             |             |             | quality and       |
|             |             |             | reputation.       |
+-------------+-------------+-------------+-------------------+
| S tatement  | sop         | 1 to 5 with | Evaluates the     |
| of Purpose  |             | 5 being the | strength and      |
| (SOP)       |             | highest     | quality of a      |
| Strength    |             | rating      | candidate's SOP   |
|             |             |             | on a scale of 1   |
|             |             |             | to 5              |
+-------------+-------------+-------------+-------------------+
| Letter of   | lor         | 1 to 5 with | Evaluates the     |
| Reccomm     |             | 5 being the | strength and      |
| endation    |             | highest     | quality of a      |
| (LOR)       |             | rating      | candidate's SOP   |
| Strength    |             |             | and LOR on a      |
|             |             |             | scale of 1 to 5   |
+-------------+-------------+-------------+-------------------+
| Under       | cgpa        | 6.8 - 9.92  | Reflects a        |
| graduate    |             |             | student's         |
| GPA         |             | (10.0       | academic          |
|             |             | scale)      | performance in    |
|             |             |             | their             |
|             |             |             | undergraduate     |
|             |             |             | studies, scored   |
|             |             |             | on a 10-point     |
|             |             |             | scale             |
+-------------+-------------+-------------+-------------------+
| Research Ex | research    | 0 or 1      | Indicates whether |
| perience    |             |             | a candidate has   |
|             |             |             | research          |
|             |             |             | experience (1) or |
|             |             |             | not (0).          |
+-------------+-------------+-------------+-------------------+
| Chance of   | chance\     | 0.34 - 0.97 | Represents the    |
| Admit       | \_of_admit  |             | likelihood of a   |
|             |             | (0 to 1     | student being     |
|             |             | scale)      | admitted,         |
|             |             |             | expressed as a    |
|             |             |             | decimal between 0 |
|             |             |             | and 1             |
+-------------+-------------+-------------+-------------------+

### Libraries

Load necessary packages for analysis and modeling.

```{r, warning=FALSE, message=FALSE}
#install.packages("janitor")
#install.packages("caTools")
#Insatll required packages
#install.packages('caret')
library(caret)
library(reshape2)
library(tidyverse)
library(ggplot2)
library(readr)
library(dplyr)
library(betareg)
library(lmtest)
library(car)
library(rcompanion)
library(janitor)
library(here)
library(caTools)
library(pROC)
```

### Loading Dataset

```{r, warning=FALSE, message=FALSE}
admission <- read_csv("adm_data.csv")
#str(admission)

admission <- admission %>%
  clean_names()

head(admission,5)
```

### Model Assumptions

+--------------+-------------------------------------+--------------+
| Assumption   | **Description**                     | Assumption   |
|              |                                     | Met          |
+==============+=====================================+==============+
| Beta         | The dependent variable is           | Met          |
| Distribution | continuous and bounded between 0    |              |
| of Dependent | and 1.                              |              |
| Variable     |                                     |              |
+--------------+-------------------------------------+--------------+
| Independent  | Observations are independent of     | Assumed      |
| Observations | each other.                         |              |
+--------------+-------------------------------------+--------------+
| Homo         | The variance of error terms are     | Assumed      |
| scedasticity | constant across levels of           |              |
| of Error     | independent variables.              |              |
| Terms        |                                     |              |
+--------------+-------------------------------------+--------------+
| Linearity of | There are linear relationship       | Met          |
| Predictors   | between logit of the expected value |              |
|              | and predictors.                     |              |
+--------------+-------------------------------------+--------------+
| No Perfect   | Predictors are not perfectly        | Met          |
| Multi        | collinear.                          |              |
| collinearity |                                     |              |
+--------------+-------------------------------------+--------------+
| No Outliers  | The data does not have influential  | Met          |
|              | outliers.                           |              |
+--------------+-------------------------------------+--------------+

```{r}
hist(admission$chance_of_admit, breaks=20, main="Histogram of Chance of Admit", xlab="Chance of Admit")

```

The dependent variable is continuous and bounded between 0 and 1.

```{r}
pairs(~ chance_of_admit + gre_score + toefl_score + university_rating + sop + lor + cgpa + research, data = admission)

```

There are linear relationship between logit of the expected value and
predictors.

```{r}
library(car)
vif(lm(chance_of_admit ~ gre_score + toefl_score + university_rating + sop + lor + cgpa + research, data = admission))

```

VIF scores are not greater than the chosen threshold of 10

```{r}
boxplot(admission$chance_of_admit, main = "Boxplot of Chance of Admit", ylab = "Chance of Admit")

```

```{r}
par(mfrow = c(2, 4))  # Adjust layout to display multiple plots
boxplot(admission$gre_score, main = "GRE Score")
boxplot(admission$toefl_score, main = "TOEFL Score")
boxplot(admission$university_rating, main = "University Rating")
boxplot(admission$sop, main = "SOP")
boxplot(admission$lor, main = "LOR")
boxplot(admission$cgpa, main = "CGPA")
boxplot(admission$research, main = "Research")

```

The data does not have influential outliers

### Further EDA

#### Chance of Admit and TOEFL

Based on exploratory data analysis, TOEFL scores appear to be associated
with a greater chance of admission, this chance of admission is further
augmented by higher university ratings, and research experience appears
to be a strong factor in increasing both TOEFL scores and the likelihood
of admission.

```{r, warning=FALSE, message=FALSE}
ggplot(admission, aes(x = toefl_score, y = chance_of_admit)) +
  geom_point(color = "#1f77b4", alpha = 0.6) +
  geom_smooth(method = "lm", color = "black", se = FALSE) +
  geom_smooth(method = "loess", color = "orange", se = FALSE) +
  labs(x = "TOEFL Score", y = "Chance of Admission", title = "Linear and Smooth Fit: Chance of Admission vs TOEFL Score") +
  theme_minimal() +
  theme(plot.title = element_text(color = "black", size = 14, face = "bold"),
        axis.title = element_text(size = 12),
        axis.text = element_text(size = 10),
        legend.position = "bottom")


```

The correlation coefficient of 0.791594 between TOEFL scores and the
chance of admit suggests a strong positive relationship. This indicates
that as TOEFL scores increase, the chance of admission tends to increase
as well.

```{r, warning=FALSE, message=FALSE}
ggplot(admission, aes(x = toefl_score, y = chance_of_admit, color = as.factor(university_rating))) +
  geom_jitter(width = .2, size = I(3)) +
  scale_color_manual(values = c("#1f77b4", "#ff7f0e", "#2ca02c", "#d62728", "#9467bd")) +
  labs(x = "TOEFL Score", y = "Chance of Admission", color = "University Rating", 
       title = "Chance of Admit by TOEFL Score per University Ranking") +
  theme_minimal() +
  theme(plot.title = element_text(color = "black", size = 14, face = "bold"),
        axis.title = element_text(size = 12),
        axis.text = element_text(size = 10)) +
  geom_smooth(method = "lm", se = FALSE)


```

This trend indicates that applicants to higher-rated universities
generally have a higher chance of admission. The standard deviation
decreases as the university rating increases, suggesting more
consistency in admission chances at higher-rated universities.

```{r, warning=FALSE, message=FALSE}
ggplot(admission, aes(x = toefl_score, y = chance_of_admit, color = as.factor(research))) +
  geom_jitter(width = .2) +
  scale_color_manual(values = c("#270181", "coral")) +
  labs(x = "TOEFL Score", y = "Chance of Admission", color = "Research Experience", 
       title = "Chance of Admit by TOEFL Score per Research Experience") +
  theme_minimal() +
  theme(plot.title = element_text(color = "black", size = 14, face = "bold"),
        legend.position = "bottom") +
  geom_smooth(method = "lm", se = FALSE)

```

Applicants with research experience (Research = 1) have a higher average
TOEFL score (approx. 110) compared to those without research experience
(Research = 0), who have an average TOEFL score of about 104.

The mean chance of admission for applicants with research experience is
significantly higher (approx. 0.796) than for those without (approx.
0.638).

This data suggests that research experience is positively associated
with both higher TOEFL scores and a greater likelihood of admission.

#### Chance of Admit and GRE

These analyses suggest that higher GRE scores are strongly correlated
with an increased chance of admission. The likelihood of admission also
appears to be influenced by the university rating and is further
enhanced by research experience.

```{r, warning=FALSE, message=FALSE}
library(ggplot2)

ggplot(admission, aes(x = gre_score, y = chance_of_admit)) +
  geom_point(color = "#1f77b4", alpha = 0.6) +
  geom_smooth(method = "lm", color = "black", se = FALSE) +
  geom_smooth(method = "loess", color = "orange", se = FALSE) +
  labs(x = "GRE Score", y = "Chance of Admission", 
       title = "Chance of Admit vs GRE Score") +
  theme_minimal() +
  theme(plot.title = element_text(color = "black", size = 14, face = "bold"),
        axis.title = element_text(size = 12),
        axis.text = element_text(size = 10),
        legend.position = "bottom") +
  guides(color = guide_legend(title = "Type of Fit", 
                              override.aes = list(linetype = c("solid", "dashed"))))


```

A correlation coefficient of 0.8026105 indicates a strong positive
relationship between GRE scores and the chance of admission. This
suggests that higher GRE scores are generally associated with a higher
likelihood of being admitted.

```{r, warning=FALSE, message=FALSE}
ggplot(admission, aes(x = gre_score, y = chance_of_admit, color = as.factor(university_rating))) +
  geom_point() + 
  scale_color_manual(values = c("#1f77b4", "#ff7f0e", "#2ca02c", "#d62728", "#9467bd")) +
  labs(x = "GRE Score", y = "Chance of Admission", color = "University Rating", 
       title = "Chance of Admit by GRE Score per University Ranking") +
  theme_minimal() +
  theme(plot.title = element_text(color = "black", size = 14, face = "bold"),
        legend.position = "bottom") +
  geom_smooth(method = "lm", se = FALSE)

```

This trend suggests that applicants to higher-rated universities have a
higher chance of admission, with the chance of admission being most
favorable at the highest-rated universities.

```{r, warning=FALSE, message=FALSE}
ggplot(admission, aes(x = gre_score, y = chance_of_admit, color = as.factor(research))) +
  geom_point() + 
  scale_color_manual(values = c("#270181", "coral")) +
  labs(x = "GRE Score", y = "Chance of Admission", color = "Research Experience", 
       title = "Chance of Admit by GRE Score per Research Experience") +
  theme_minimal() +
  theme(plot.title = element_text(color = "black", size = 14, face = "bold"),
        legend.position = "bottom") +
  geom_smooth(method = "lm", se = FALSE)

```

Applicants with research experience (Research = 1) have a higher average
GRE score (about 323) compared to those without research experience
(Research = 0), who have an average GRE score of approximately 309.
Similarly, the mean chance of admission is significantly higher for
applicants with research experience (approx. 0.796) than for those
without it (approx. 0.638). This indicates that research experience is
positively associated with both higher GRE scores and a greater
likelihood of admission.

#### Chance of Admit and CGPA

```{r, warning=FALSE, message=FALSE}
ggplot(admission, aes(x = cgpa, y = chance_of_admit, color = as.factor(university_rating))) +
  geom_point() + 
  scale_color_manual(values = c("#1f77b4", "#ff7f0e", "#2ca02c", "#d62728", "#9467bd")) +
  labs(x = "G.P.A Score", y = "Chance of Admission", color = "University Rating", 
       title = "Chance of Admit by G.P.A per University Ranking") +
  theme_minimal() +
  theme(plot.title = element_text(color = "black", size = 14, face = "bold"),
        legend.position = "bottom") +
  geom_smooth(method = "lm", se = FALSE)
```

Applicants with a higher g.p.a have a higher acceptance probability as
the university ranking goes from low(1) to high(5). The correlation
value of 0.87 indicates a strong positive relationship between G.P.A
score and the chance of admission. This suggests that higher a G.P.A
score are generally associated with a higher likelihood of being
admitted.

#### Chance of Admission Correlation Heatmap

```{r, warning=FALSE, message=FALSE}
library(ggplot2)
library(reshape2)
library(viridis)

# Calculate the correlation matrix
data <- cor(admission[sapply(admission, is.numeric)], use = "complete.obs")

# Reshape data for ggplot
data1 <- melt(data)

# Create the heatmap
ggplot(data1, aes(x = Var1, y = Var2, fill = value)) +
  geom_tile() +
  scale_fill_viridis(option = "C", direction = -1) +  # Using viridis color scale
  labs(title = "Admission Correlation Heatmap", x = "", y = "") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1, size = 10),
        axis.text.y = element_text(size = 10),
        plot.title = element_text(color = "black", size = 14, face = "bold"),
        legend.position = "right") +
  geom_text(aes(label = round(value, digits = 2)), color = "white", size = 3)

  
```

The heatmap shows that GRE scores, TOEFL scores, and CGPA are strongly
and positively correlated with the chance of admission. Research
experience also positively influences admission chances, albeit to a
lesser extent than academic scores.

## Analysis and Results

Given the exploratory data analysis, we will now construct a predictive
model for the response variable chance_of_admit.

```{r}
# Scaling of Data
#admission <- scale(admission)
#admission <- as.data.frame(admission)
```

```{r}
# Splitting the Data

#make this example reproducible
set.seed(1)

#Use 70% of dataset as training set and remaining 30% as testing set
sample <- sample(c(TRUE, FALSE), nrow(admission), replace=TRUE, prob=c(0.7,0.3))
train  <- admission[sample, ]
test   <- admission[!sample, ]

#view dimensions of training set
#dim(train)

#view dimensions of test set
#dim(test)

X_train <- train

X_test = subset(test,select = -c(chance_of_admit))
keeps <- c("chance_of_admit")
y_test = test[keeps]
```

## Fitting Full Model

Model 1 = chance_of_admit \~ gre_score + toefl_score +
university_rating + sop + lor + cgpa + research

```{r}
gy_logit <- betareg(chance_of_admit ~ gre_score + toefl_score + university_rating + sop + lor + cgpa + research, data = train)

model_summary <- summary(gy_logit)
# Extracting coefficients for the mean model
coefficients_mean <- model_summary$coefficients$mean
coeff_mean_df <- data.frame(round(coefficients_mean, 4))
colnames(coeff_mean_df) <- c("Estimate", "Std. Error", "z value", "Pr(>|z|)")

# Extract log-likelihood and pseudo R-squared
log_likelihood <- round(model_summary$loglik, 2)
pseudo_r_squared <- round(model_summary$pseudo.r.squared, 4)

# Print the formatted results
print("Coefficients (Mean Model):")
print(coeff_mean_df)
cat("\nLog-Likelihood:", log_likelihood, "\nPseudo R-squared:", pseudo_r_squared)


coefficients_df <- data.frame(
  Term = rownames(model_summary$coefficients$mean),
  Estimate = model_summary$coefficients$mean[, "Estimate"],
  Std_Error = model_summary$coefficients$mean[, "Std. Error"]
)
coefficients_df <- coefficients_df[coefficients_df$Term != "(Intercept)",]

ggplot(coefficients_df, aes(x = Term, y = Estimate)) +
  geom_point() +
  geom_errorbar(aes(ymin = Estimate - Std_Error, ymax = Estimate + Std_Error), width = 0.2) +
  theme_minimal() +
  coord_flip() + 
  xlab("Model Terms") +
  ylab("Coefficient Estimate") +
  ggtitle("Coefficient Estimates with Standard Errors")

```

Next we use the'regsubsets' function to perform subset selection, aiming
to identify the most predictive variables for the 'chance_of_admit'
outcome. This approach systematically evaluates combinations of up to
seven predictors, such as 'gre_score', 'toefl_score', 'cgpa', and
others, to determine their impact on the chance of admission. The goal
is to find the optimal set of variables that best predict the admission
outcome with varying model complexities.

The output indicates which variables are included in the best models at
each complexity level, revealing that 'cgpa', 'gre_score', and 'lor' are
significant predictors in simpler models. As more variables are added,
the model becomes more complex, potentially improving accuracy but also
increasing the risk of overfitting. This analysis helps in understanding
the relative importance of different academic and profile factors in
determining admission chances. In the analyzing the predictors and their
impact, it is determined that the greatest factor affecting the
probability of admission is g.p.a. by a factor of.64. This is followed
by research with a factor of 0.15 and the combination of statement of
purpose and letter of recommendation with a factor of 0.07.

```{r}
library(leaps)
models <- regsubsets(chance_of_admit~. -serial_no, data = admission, nvmax = 9,
                     method = "seqrep")

library(broom)

model_summaries <- summary(models)

# Number of models
num_models <- nrow(model_summaries$which)

# Initialize an empty data frame for the summary
summary_table <- data.frame(Size = integer(), 
                            Variables = character(), 
                            R.Squared = numeric(), 
                            Adj.R.Squared = numeric(), 
                            BIC = numeric(), 
                            stringsAsFactors = FALSE)

# Fill the summary table
for (i in 1:num_models) {
  included_vars <- names(which(model_summaries$which[i, ]))
  vars_str <- paste(included_vars, collapse = ", ")
  summary_table <- rbind(summary_table, 
                         data.frame(Size = i, 
                                    Variables = vars_str, 
                                    R.Squared = model_summaries$rsq[i], 
                                    Adj.R.Squared = model_summaries$adjr2[i], 
                                    BIC = model_summaries$bic[i]))
}

# Print the table
print(summary_table)



```

In this analysis 10-fold cross-validation is employed to provide
assessment of the model's predictive performance by dividing the data
into ten subsets and iteratively training the model on nine subsets
while testing on the remaining one. The procedure is repeated for models
with varying numbers of predictors. Key performance metrics such as RMSE
(Root Mean Squared Error), R-squared, and MAE (Mean Absolute Error) are
calculated for each model, showing how the inclusion of additional
variables impacts the model's accuracy. The results show variations in
model performance across different numbers of predictors, with certain
models achieving lower RMSE and higher R-squared values

```{r}
# Set seed for reproducibility
set.seed(123)
# Set up repeated k-fold cross-validation
train.control <- trainControl(method = "cv", number = 10)
# Train the model
step.model <- train(chance_of_admit ~. -serial_no, data = admission,
                    method = "leapSeq", 
                    tuneGrid = data.frame(nvmax = 1:7),
                    trControl = train.control
                    )
step.model$results
```

We employ beta regression models to predict the chance of admission,
utilizing various combinations of predictors such as GRE scores, TOEFL
scores, university ratings, letters of recommendation, CGPA, and
research experience. The primary objective is to determine the optimal
set of predictors that can accurately forecast admission chances while
maintaining model simplicity.

The models, labeled 'gy_logit' to 'gy_logit7', were evaluated using the
Akaike Information Criterion (AIC) and the pseudo R-squared metric. A
lower AIC value indicates a more efficient model in terms of the
trade-off between goodness of fit and complexity, while a higher pseudo
R-squared value suggests a better model fit.

The outcomes reveal that 'gy_logit4' and 'gy_logit7' yield the lowest
AIC values, indicating they are potentially the most efficient models
among those tested. Both these models, along with 'gy_logit', also
exhibit the highest pseudo R-squared values, suggesting strong model
fits.

```{r}
gy_logit <- betareg(chance_of_admit ~ gre_score + toefl_score + university_rating + lor + cgpa + research, data = train)

gy_logit1 <- betareg(chance_of_admit ~ gre_score + toefl_score + university_rating + lor + cgpa , data = train)

gy_logit2 <- betareg(chance_of_admit ~ gre_score + toefl_score + university_rating + lor + research, data = train)

gy_logit3 <- betareg(chance_of_admit ~ gre_score + toefl_score + university_rating + cgpa + research, data = train)

gy_logit4 <- betareg(chance_of_admit ~ gre_score + toefl_score + lor + cgpa + research, data = train)

gy_logit5 <- betareg(chance_of_admit ~ gre_score + university_rating + lor + cgpa + research, data = train)

gy_logit6 <- betareg(chance_of_admit ~ toefl_score + university_rating + lor + cgpa + research, data = train)

gy_logit7 <- betareg(chance_of_admit ~ gre_score + toefl_score + university_rating + lor + cgpa + research, data = train)

# Creating a data frame with AIC and pseudo R-squared values for each model
model_comparison <- data.frame(
  Model = c("gy_logit", "gy_logit1", "gy_logit2", "gy_logit3", "gy_logit4", "gy_logit5", "gy_logit6", "gy_logit7"),
  AIC = c(AIC(gy_logit), AIC(gy_logit1), AIC(gy_logit2), AIC(gy_logit3), AIC(gy_logit4), AIC(gy_logit5), AIC(gy_logit6), AIC(gy_logit7)),
  Pseudo_R_Squared = c(gy_logit$pseudo.r.squared, gy_logit1$pseudo.r.squared, gy_logit2$pseudo.r.squared, gy_logit3$pseudo.r.squared, gy_logit4$pseudo.r.squared, gy_logit5$pseudo.r.squared, gy_logit6$pseudo.r.squared, gy_logit7$pseudo.r.squared)
)


print(model_comparison)

```

### Conclusion

In concluding our study, the application of beta regression models has
provided valuable insights into predicting student admission
probabilities. The analysis highlights the significance of various
factors, such as GRE scores, TOEFL scores, CGPA, university ratings,
letters of recommendation, and research experience. Practically
speaking, GRE and TOEFL scores are critical as they are standard
measures of a student's academic readiness and language proficiency,
which are crucial for success in higher education settings. CGPA
reflects consistent academic performance, while university ratings may
correlate with the perceived quality and competitiveness of applicants.
Letters of recommendation and research experience offer a qualitative
assessment of a student's capabilities and potential contributions to
academic discourse.

The models that combined these predictors with the lowest AIC values and
the highest pseudo R-squared scores were deemed most effective. They
balance predictive accuracy with model simplicity, avoiding the pitfalls
of overfitting associated with more complex models. This balance is
vital for practical application, ensuring that the model remains
generalizable and relevant to various educational contexts.

Looking forward, there are several avenues for further research and
development. One potential area is the exploration of additional
variables that could impact admission chances, such as extracurricular
activities, personal statements, or socio-economic background. Another
aspect worth investigating is the application of these models in
different educational contexts and geographical locations to assess
their universality and adaptability.

Additionally, further refinement of the models could involve exploring
alternative statistical techniques or more complex machine learning
algorithms that might capture nonlinear relationships and interactions
more effectively. It would also be beneficial to consider the ethical
implications of using such predictive models in admission processes,
ensuring fairness and diversity in student selection.

In summary, this study not only contributes to the academic
understanding of factors influencing university admissions but also
offers practical implications for educational institutions and
policy-making. By leveraging statistical modeling, universities can gain
a more nuanced understanding of the admission process, aiding in the
development of more informed and equitable admission policies.

## References
